Co-evolution of lexical meaning \& pragmatic use

### 2018-03-30 paper/cogsci-submission2.tex
final manuscript+letter changes TB

In contrast, we will here represent explicitly language users' representations of lexical
meanings and their own particular manner of manner of production and interpretation based on
these lexical representation. Different types of pragmatic behavior are represented using
probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}.

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favors communicative efficiency, languages are driven towards semantic
expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But pressure towards
communicative efficiency is not the only force that shapes language. Learnability is another.
Natural languages need to be learnable to survive their faithful transmission across
generations. Furthermore, the effects of even small learning biases can have striking effects
on an evolving language in a process of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both communicative efficiency and learnability, these forces may
pull in opposite directions \citep[\S7]{christiansen+chater:2008}. Their opposition becomes particularly clear when considering the
extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
form-meaning association is easy to learn but likely bad for communication with peers. Conversely, a language
that lexicalizes distinct forms for a large number of different meanings may be better for communication 
but challenging to acquire.

@@

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent reproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of $f_j$, i.e., of how well $j$ fares against the
population: the better the behavior $j$, the more likely it is to be imitated.\footnote{A
  crucial assumption in all of the derivations of the replicator dynamic as a biological or
  cultural process of evolution is that agents themselves are rather limited in their strategic
  capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
  type, that is optimal given the current population state. They are also not able to
  strategically adapt their behavior to any particular agent they are interacting with on some
  given occasion. Such minimalism about the strategic capacities of agents is a conservative
  methodological choice. It is one thing to be able to demonstrate that phenomenon $X$ is what
  a rational agent would do. This is useful but may not always be the most satisfactory answer
  to the question how $X$ happened to evolve. It is another thing to show that even if no agent
  was individually capable of seeing $X$ as the rational thing to do, $X$ would still evolve if
  every agent at least sometimes, locally and myopically tried to optimize its behavior, e.g.,
  by imitating what seems to be working well for others \citep{Skyrms2010:Signals}.}

@@

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution promotes or demotes. A type determines an agent's behavior and thereby
an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
the joint evolution of semantic meaning and pragmatic use, types are pairs consisting of a
lexicon and a linguistic strategy of language use.

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use
Bayes' rule to weigh interpretations based on a conjecture about speaker behavior. Speaker
behavior is regulated by a soft-max parameter $\lambda \geq 0$
\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices approximate strict
maximization of expected utilities. Expected utility of a message $\messg$ in state $\state$
for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer
will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
  utility function, frequently made in Bayesian approaches after \citet{frank+goodman:2012}, is
  $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the logarithm, utilities
  are derived from action choices, while with the logarithmic utilities are derived from how
  closely the listener's beliefs (after hearing an utterance) align with the speaker's, as
  measure by Kullback-Leibler divergence. In an evolutionary context, where fitness comes from
  actual interactions with the environment, not just solemn beliefs held about it, the
  formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
  utilities cause the speaker never to use false utterances. Since we aim here for a situation
  in which any type has a reasonable chance to be mistaken for any other based on their
  production behavior (see Section~\ref{sec:learnability}), we want speaker types which send
  any message in any state with positive probability.} For literal speakers, utility only
tracks truthfulness. Literal speakers choose any true message with equal probability but may
send false messages as well with a probability dependent on $\lambda$.

In words, an agent's linguistic behavior -- what message to send when or which message to
interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows,
this is what we identify an agent's type with, in order to analyze under which conditions
combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior
(level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend
solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior
(level-$n+1$ reasoning) results from reasoning about rational language use. Such agents produce
and comprehend by reasoning about how they themselves would interpret or use
expressions. Higher order behavior is therefore, even if in tendency rational, quite simple. It
does not assume agents to know their interlocutor's actual type when interacting (e.g., what
lexicon the interlocutor is using); they simply act based on a conjecture about language use
derived from their own (solipsistic) perspective.

@@

With low $\lambda$ speakers choose true messages with more slack. Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because they reason that $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$. For $\lambda = 20$, there will be less slack in literal speaker behavior:

@@

When it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at
level $1$ is not advantageous. The reason for this is that literal use of $\Lbound$ already
endows agents with a behavioral strategy that associates a single state with a single message
(in tendency; depending on $\lambda$ for senders). For $\Lbound$-receivers of level $1$,
reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally
decrease the association of one state with one message. This decrease is only slight if
$\lambda$ is high, but nevertheless present. That is to say, level-$1$ reasoning does not
necessarily confer a functional advantage. For some types, such as users of $\Lbound$, literal
signaling is preferable. In sum, the definition of types introduced here constitute conditions
that are purposefully averse to what we would like to show; the evolution of is not immediate
just based on fitness-based selection, to which we turn next.

@@

Communicative success and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the update step, its fitness $f_i$, and the probability that a learner
infers $i$ when observing the behavior of a type-$j$ agent. Fitness-based can be thought of as
biological (fitness as expected relative number of offspring) or cultural evolution (fitness as
likelihood of being imitated or repeated, see Appendix~\ref{sec:deriv-repl-mutat} for a
concrete example). A type's communicative success depends on how well it communicates within a
population while its learnability depends on the fidelity by which it is inferred by new
generations of learners. The learners' task is consequently to perform a joint inference over
types of linguistic behavior and lexical meaning.

@@

\begin{figure}
\centering
    % \begin{tikzpicture}

  %   \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);

  %   \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
  %   \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
  %   \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
  %   \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};

  %   \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
  %   \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
  %   \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
  %   \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};

  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.75,2.45);
  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.25,2.75);
  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.75,2.75);

  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{biased}};
  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};

  %   \end{tikzpicture}
\caption{Sketch of dynamics on a two-dimensional type space with four types. The $y$-axis
  represents the proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows
  sketch population trajectories under functional pressure (replicator steps, red arrows) and
  learnability pressure (mutator steps) with either a flat prior (blue dash-dotted arrows) or a
  prior that favors $\Llack$ (solid purple arrows).\mf{I did not change the upper-most arrow
    for biased learnability, because I cannot compile this anyway; I did change some typos in
    arrow labelling though.}}
\label{fig:sketch}
\end{figure}

This section illustrates how replication and mutation act on the evolution of semantic meaning
and pragmatic use in the simplest non-trivial type space, which consists of only the four types
from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and it is
either a literal (level-$0$) or a pragmatic (level-$1$) language user. We may think of such a
population as a point in a two-dimensional space. The grey dotted square in
Fig.~\ref{fig:sketch} outlines this space. The space's first dimension determines the
proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
Fig.~\ref{fig:sketch}). The second dimension determines the proportion of pragmatic language
users (the $x$-axis in Fig.~\ref{fig:sketch}). Populations at the corners of this space are
monomorphic; all agents in them are of a single type. Points away from the corners represent
mixed populations of various types.\footnote{This two-dimensional representation underspecifies
  the exact proportions of types found in a mixed population. The space's center, for instance,
  could be composed in equal parts of pragmatic $\Llack$ and of literal $\Lbound$. But it could
  equally well be a population that has all four types in equal proportions. This level of
  granularity will nevertheless be sufficient for our purposes, as we are only interested in
  sketching rough trajectories after replicator or mutator steps.} We can predict general
trends in evolutionary trajectories in this space based on what we know from the literature and
the setup from Section~\ref{sec:fitn--learn}, even before any further simulation results.

When it comes to communicative success, types with $\Lbound$ have a functional advantage over
$\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch trajectories we
can expect if there is only pressure for communicative success. How much each replicator step
moves a population through this space ultimately depends on the rationality parameter
$\lambda$. Functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently
high, all types except for literal $\Llack$ speakers will show a strong tendency to associate
$\ssome$ with $\msome$ and $\sall$ with $\mall$.

Under probability matching $l=1$, iterated Bayesian learning converges to the prior over types
\citep{griffiths+kalish:2007}. If the prior on types is uniform, populations gravitate toward
the space's center under repeated mutator steps. This is sketched by the blue dash-dotted
arrows in Fig.~\ref{fig:sketch}. If instead the prior favors $\Llack$ over $\Lbound$, we expect
repeated mutator steps to drive the population toward the upper-half of the space. This is
suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. The system's predictions are
more difficult to predict when $l>1$, because now individual mutator steps may also be
influenced not only by the priors, but also by the differential likelihood with which a
particular type $i$ can be confused for any other type $j$ (see Section~\ref{sec:learnability}
and Appendix~\ref{sec:example-iter-bayes}).

Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
convention will not evolve without disposition towards pragmatic enrichments (left part of
Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
enrichment can co-evolve. These abstract theoretical considerations also suggest that we need
both fitness-based selection of communicative efficiency and pressure towards learnability to
see underspecified semantics and pragmatic language use emerge together.

\begin{figure}[t]
\centering
\includegraphics[width=1.08\textwidth,height=10cm, keepaspectratio]{./plots/fig-contourb105k5.png}
\caption{Dynamics on a two-dimensional type space with four types. The $y$-axis represents the
  proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows indicate
  directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior
  that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$. Colored
  contours show the proportion of pragmatic $\Llack$ users after a step. \tb{I don't think the
    contours are adding much, so they could be removed. Then again, I think they are rather
    non-intrusive, look nice, and illustrate how the step-size, depending on where we are in
    the space, varies.} \mf{I think that the contours are sweet. But I worried about the size
  of the arrows. I made the picture a bit bigger}}

@@

Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
$\Llack$ against $\Lbound$ apparent. With increasing $\lambda$, this difference pragmatic users
of $\Lbound$ are no longer selected against by fitness-based selection.

>>>>

In contrast, we will here explicitly model language users' representations of lexical
meanings and their own particular manner of production and interpretation based on
these lexical representations. Different types of pragmatic behavior -- ways of deploying meaning in interaction -- are represented using
probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}.

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor communicative efficiency, languages are driven towards semantic
expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But pressure towards
communicative efficiency is not the only force that shapes language. Learnability is another.
Natural languages need to be learnable to survive their faithful transmission across
generations. Furthermore, the effects of even small learning biases can have striking effects
on an evolving language in a process of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both communicative efficiency and learnability, these forces may
pull in opposite directions \citep[\S7]{christiansen+chater:2008}. Their opposition becomes particularly clear when considering the
extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
form-meaning association is easy to learn but likely bad for communication with peers. Conversely, a language
that lexicalizes distinct forms for a large number of different meanings may be better for communication 
but more challenging to acquire.

@@

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent reproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of $f_j$, i.e., of how well $j$ fares against the
population: the better the behavior $j$, the more likely it is to be imitated.\footnote{A
  crucial assumption in all of the derivations of the replicator dynamic as a biological or
  cultural process of evolution is that agents themselves are rather limited in their strategic
  capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
  type, that is optimal given the current population state. They are also not able to
  strategically adapt their behavior to any particular agent they are interacting with on some
  given occasion. Such minimalism about the strategic capacities of agents is a conservative
  methodological choice. It is one thing to be able to demonstrate that phenomenon $X$ is what
  a rational agent would do. This is useful but may not always be the most satisfactory answer
  to the question how $X$ happened to evolve. It is another thing to show that even if no agent
  was individually capable of seeing $X$ as the rational thing to do, $X$ would still evolve if
  every agent at least sometimes, locally and myopically, tried to optimize its behavior, e.g.,
  by imitating what seems to be working well for others \citep{Skyrms2010:Signals}.}

@@

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution promotes or demotes. Its type determines an agent's behavior and thereby
an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
the joint evolution of semantic meaning and pragmatic use, types are pairs consisting of a
lexicon and a linguistic strategy of language use.

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is in the following assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use
Bayes' rule to weigh interpretations based on a conjecture about speaker behavior. Speaker
behavior is regulated by a soft-max parameter $\lambda \geq 0$
\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices approximate strict
maximization of expected utilities. Expected utility of a message $\messg$ in state $\state$
for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer
will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
  utility function, frequently made in Bayesian approaches after \citet{frank+goodman:2012}, is
  $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the logarithm, utilities
  are derived from action choices, while with the logarithmic utilities are derived from how
  closely the listener's beliefs (after hearing an utterance) align with the speaker's, as
  measure by Kullback-Leibler divergence. In an evolutionary context, where fitness comes from
  actual interactions with the environment, not just solemn beliefs held about it, the
  formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
  utilities cause the speaker never to use false utterances. Since we aim here for a situation
  in which any type has a reasonable chance to be mistaken for any other based on their
  production behavior (see Section~\ref{sec:learnability}), we want speaker types which send
  any message in any state with positive probability.} For literal speakers, utility only
tracks truthfulness. Literal speakers choose any true message with equal probability but may
send false messages as well with a probability dependent on $\lambda$.

In words, an agent's linguistic behavior -- what message to send when or which message to
interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows,
this is what we identify an agent's type with, in order to analyze under which conditions
combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior
(level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend
solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior
(level-$n+1$ reasoning) results from reasoning about rational language use. Such agents produce
and comprehend by reasoning about how they themselves would interpret or use
expressions. Higher order behavior is therefore, even if in tendency rational, quite simple. It
does not assume agents to know their interlocutor's type when interacting (e.g., what
lexicon the interlocutor is using); they simply act based on a conjecture about language use
derived from their own (solipsistic) perspective.

@@

With low $\lambda$ speakers choose true messages with more slack. Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users, because they reason that $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$. For $\lambda = 20$, there will be less slack in literal speaker behavior:

@@

When it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at
level $1$ is not advantageous. The reason for this is that literal use of $\Lbound$ already
endows agents with a behavioral strategy that associates a single state with a single message
(in tendency; depending on $\lambda$ for senders). For $\Lbound$-receivers of level $1$,
reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally
decrease the association of one state with one message. This decrease is only slight if
$\lambda$ is high, but nevertheless present. That is to say, level-$1$ reasoning does not
necessarily confer a functional advantage. For some types, such as users of $\Lbound$, literal
signaling is preferable. In sum, the definition of types introduced here constitute conditions
that are purposefully averse to what we would like to show; the evolution of a division of labor between semantics and pragmatics is not immediate
just based on fitness-based selection, to which we turn next.

@@

Communicative success and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the update step, its fitness $f_i$, and the probability that a learner
infers $i$ when observing the behavior of a type-$j$ agent. Fitness-based replication can be thought of as
biological (fitness as expected relative number of offspring) or cultural evolution (fitness as
likelihood of being imitated or repeated, see Appendix~\ref{sec:deriv-repl-mutat} for a
concrete example). A type's communicative success depends on how well it communicates within its 
population while its learnability depends on the fidelity by which it is inferred by new
generations of learners. The learners' task is consequently to perform a joint inference over
types of linguistic behavior and lexical meaning.

@@

\begin{figure}
\centering
     \begin{tikzpicture}

     \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);

     \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
     \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
     \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
     \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};

     \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
     \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
     \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
     \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};

   \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
   \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.75,2.45);
   \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.25,2.75);
   \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.75,2.75);

     \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
     \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
     \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
     \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{biased}};
     \draw[->, thick, goldish] (0.75,5.25) -- (2.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
     \draw[<-, thick, goldish] (2.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};

     \end{tikzpicture}
\caption{Sketch of dynamics on a two-dimensional type space with four types. The $y$-axis
  represents the proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows
  sketch population trajectories under functional pressure (replicator steps, red arrows) and
  learnability pressure (mutator steps) with either a flat prior (blue dash-dotted arrows) or a
  prior that favors $\Llack$ (solid purple arrows).}
\label{fig:sketch}
\end{figure}

This section illustrates how replication and mutation act on the evolution of semantic meaning
and pragmatic use in the simplest non-trivial type space, which consists of only the four types
from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and it is
either a literal (level-$0$) or a pragmatic (level-$1$) language user. We may think of such a
population as a point in a two-dimensional space. The grey dotted square in
Fig.~\ref{fig:sketch} outlines this space. The space's first dimension determines the
proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
Fig.~\ref{fig:sketch}). The second dimension determines the proportion of pragmatic language
users (the $x$-axis in Fig.~\ref{fig:sketch}). Populations at the corners of this space are
monomorphic; all agents in them are of a single type. Points away from the corners represent
mixed populations of various types.\footnote{This two-dimensional representation underspecifies
  the exact proportions of types found in a mixed population. The space's center, for instance,
  could be composed in equal parts of pragmatic $\Llack$ and of literal $\Lbound$. But it could
  equally well be a population that has all four types in equal proportions. This level of
  granularity will nevertheless be sufficient for our purposes, as we are only interested in
  sketching rough trajectories after replicator or mutator steps.} We can predict general
trends in evolutionary trajectories in this space based on what we know from the literature and
the setup from Section~\ref{sec:fitn--learn}, even before any simulation results.

When it comes to communicative success, types with $\Lbound$ have a functional advantage over
$\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch trajectories we
can accordingly expect if there is only pressure for communicative success. How much each replicator step
moves a population through this space ultimately depends on the rationality parameter
$\lambda$. Functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently
high, all types except for literal $\Llack$ speakers will show a strong tendency to associate
$\ssome$ with $\msome$ and $\sall$ with $\mall$.

Under probability matching, $l=1$, iterated Bayesian learning converges to the prior over types
\citep{griffiths+kalish:2007}. If the prior on types is uniform, populations gravitate toward
the space's center under repeated mutator steps. This is sketched by the blue dash-dotted
arrows in Fig.~\ref{fig:sketch}. If instead the prior favors $\Llack$ over $\Lbound$, we expect
repeated mutator steps to drive the population toward the upper-half of the space. This is
suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. The system's predictions are
more difficult to predict when $l>1$, because now individual mutator steps may also be
influenced not only by the prior, but also by the differential likelihood with which a
particular type $i$ can be confused for any other type $j$ (see Section~\ref{sec:learnability}
and Appendix~\ref{sec:example-iter-bayes}).

Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under strong
and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
convention will not evolve without disposition towards pragmatic enrichments (left part of
Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
enrichment can co-evolve. These abstract theoretical considerations also suggest that we need
both fitness-based selection of communicative efficiency and pressure towards learnability to
see underspecified semantics and pragmatic language use emerge together.

\begin{figure}[t]
\centering
\includegraphics[width=1.08\textwidth,height=10cm, keepaspectratio]{./plots/fig-contourb105k5.png}
\caption{Dynamics on a two-dimensional type space with four types. The $y$-axis represents the
  proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows indicate
  directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior
  that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$. Colored
  contours show the proportion of pragmatic $\Llack$ users after a step.}

@@

Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
$\Llack$ against $\Lbound$ apparent. With increasing $\lambda$, this difference is mitigated for pragmatic users of $\Lbound$, who are no longer selected against by fitness-based selection.
